Quotients of the order complex $\Delta(\overline{\Pi}_n)$ by subgroups of the Young subgroup $S_1\times S_{n-1}$

### 2018-05-17 paper1/paper.tex
mistakes in paper1 fixed

In this paper we give an answer for all Young subgroups that fix the element $1$, which are of the form $S_1\times S_{k_1}\times\dots\times S_{k_r}$, as a corollary of the more general Proposition \ref{rdthm}. An instance, which we have not yet solved completely, are Young subgroups of the form $S_k\times S_{n-k}$. However, we have computed the first homology with integer coefficients, obtaining the result $H_1(\Delta(\overline{\Pi}_n)/S_k\times S_{n-k};\Z)\cong 0$ for $n\geq 5$.
\section{Discrete Morse Theory}
The definitions of regular trisps, partial matchings, acyclic matchings and foundations of Discrete Morse Theory can be found in \cite{forman}, \cite{clmap} and \cite{buch}. The following two theorems of Discrete Morse Theory are frequently used in our proofs. We also make use of the results in \cite{clmap}.
\begin{thm}[Patchwork Theorem]
\label{patchwork}
Let $\varphi:P\longrightarrow Q$ be an order-preserving map and assume we have acyclic matchings on the subposets $\varphi^{-1}(q)$ for all $q\in Q$. Then the union of these matchings is an acyclic matching on $P$.
\end{thm}
\begin{thm}%[Main Theorem of Discrete Morse Theory for CW Complexes]
\label{morsemain}
Let $\Delta$ be a regular trisp and let $M$ be an acyclic matching on the poset ${\cal F}(\Delta)\setminus\{\hat{0}\}$. Let $c_i$ denote the number of critical $i$-dimensional simplices of $\Delta$. Then:
\begin{enumerate}
\item If the critical simplices form a subcomplex $\Delta_c$ of $\Delta$, then $\Delta$ collapses onto $\Delta_c$.
\item $\Delta$ is homotopy equivalent to a CW complex with $c_i$ cells of dimension $i$.
\end{enumerate}
\end{thm}
The proofs of Theorem \ref{patchwork} and \ref{morsemain} as well as further facts on Discrete Morse Theory can be found in \cite[Chapter 11]{buch}.
\section{The collapsible subtrisp}
\label{allgemein}
Let $n\geq 3$ be a fixed natural number throughout this section. The symmetric group $S_n$ operates on $\Delta(\overline{\Pi}_n)$ in a natural way. In this paper we consider subgroups of the Young subgroup
\[S_1\times S_{n-1}:=\{\sigma\in S_n\mid\sigma(1)=1\}\]
We now consider the regular trisp $\Delta:=\Delta(\overline{\Pi}_n)/G$, where $G\subset S_1\times S_{n-1}$ is a subgroup. Let $\Delta^*$ be that subtrisp of $\Delta$ that we obtain by removing all vertices that are represented by some partition where the block containing $1$ has exactly two elements and any other block is singleton. This condition is independent of the choice of the representative. Such a representative can be written as
\[
v_k:=\{\{1,k\},\{2\},\dots,\widehat{\{k\}},\dots,\{n\}\}
\]
with $k\in\{2,\ldots, n\}$. The element with the hat above is omitted.

@@

Let $B$ be the set of all vertices that are not contained in $R$. Now we define a closure map $\varphi:B\longrightarrow R$. Let $b\in B$ and let $\tilde b\in\overline{\Pi}_n$ be a representative of $b$. We modify the blocks in $\tilde b$ as follows. We replace the block $A$ containing $1$ by the two blocks $\{1\}$ and $A\setminus\{1\}$ and take the class of this new partition as $\varphi(b)$. By applying Corollary 2.5 in \cite{clmap} and Proposition 7 in \cite{jule} it is easy to see that this map is well-defined and a closure map\footnote{see Definition 2.1 in \cite{clmap}}. By Theorem 2.2 in \cite{clmap} we get an acyclic matching on ${\cal F}(\Delta^*)$ where a simplex $\sigma$ is critical if and only if  $V(\sigma)\subset R$.

@@

What about the remaining simplices that are represented by a chain that contains a $v_k$ with $k\in\{2,\ldots, n\}$? Let $V_G$ be the set of all these special vertices. We define a Poset $P_G:=V_G\cup\{0\}$ such that $0$ is the smallest element of $P_G$ and the only element that is comparable with some other element. That means $x,y\in P_G$, $x<y$ implies $x=0$. We define an order-preserving map $\varphi_G:{\cal F}(\Delta)\longrightarrow P_G$ as follows. Let $\sigma\in{\cal F}(\Delta)$, then we map $\sigma$ to $0$ if $\sigma\in{\cal F}(\Delta^*)$. Otherwise we map $\sigma$ to the special vertex of $V_G$ that belongs to $\sigma$, which is unique. By Lemma \ref{mitte} we already have an acyclic matching on $\varphi_G^{-1}(0)$ with one critical cell which has dimension $0$. The next step would be to find acyclic matchings on $\varphi_G^{-1}(v)$ for all $v\in V_G$ and apply Theorem \ref{patchwork}.

@@

as follows. Let $\sigma\in{\cal F}(\Delta(\overline{\Pi}_{n-1})/i_k^{-1}(G))$ be a simplex, then we choose a representative $\tilde\sigma\in{\cal F}(\Delta(\overline{\Pi}_{n-1}))$ and describe what happens to the vertices of $\tilde\sigma$, which are partitions of the set $[n-1]$. Let $p=\{B_1,B_2,\dots,B_r\}$, where $B_1$ is the block that contains $1$, be a partition in the chain $\tilde\sigma$. For $p$ we define $p^*:=\{f_k(B_1)\cup\{k\},f_k(B_2),\dots,f_k(B_r)\}$, which is a partition of set set $[n]$. $\tilde\sigma$ can be written as

@@

By the discussion at the end of section \ref{allgemein} it remains to show that there exists an acyclic matching on $\varphi_G^{-1}(v)$ for each $v\in V_G$ such that all critical simplices have dimension $n-3$. Let $v\in V_G$, then we choose a $2\leq k\leq n$ such that $v_k$ is a representative of $v$. By induction hypothesis we have an acyclic matching on $\Delta(\overline{\Pi}_{n-1})/i_k^{-1}(G)$ which has one critical simplex of dimension $0$ and critical simplices of dimension $n-4$. We apply Lemma \ref{indlemma} and get our acyclic matching on $\varphi_G^{-1}(v)$.
\end{proof}
\section{Counting spheres}
\label{giacomo}
We are now interested in the number of spheres we have. This can be computed either by looking into the proof of Proposition \ref{rdthm} or by determining the Betti Number $\beta_{n-3}(\Delta(\overline{\Pi}_n)/G)$, which is the dimension of $H_{n-3}(\Delta(\overline{\Pi}_n)/G;\C)$.

The action of $S_1\times S_{n-1}$ on $\Delta(\overline{\Pi}_n)$ induces an action on $H_{n-3}(\Delta(\overline{\Pi}_n);\C)$. By applying Corollary 3.3 in \cite{emk} and Theorem 1.5 in \cite{white} we know that $H_{n-3}(\Delta(\overline{\Pi}_n);\C)$ and
\[\C\cdot(S_1\times S_{n-1})=\bigoplus_{\sigma\in S_1\times S_{n-1}}\C\cdot\sigma\]
are isomorphic as $(S_1\times S_{n-1})$-modules. The Transfer Theorem tells us
\[
H_{n-3}(\Delta(\overline{\Pi}_n)/G;\C)\cong H_{n-3}(\Delta(\overline{\Pi}_n);\C)^G\cong(\C\cdot(S_1\times S_{n-1}))^G
\]
The dimension of $(\C\cdot(S_1\times S_{n-1}))^G$ is the index of $G$ in $S_1\times S_{n-1}$. Hence we obtain the following result: $\Delta(\overline{\Pi}_n)/G$ is homotopy equivalent to a wedge of $k$ spheres of dimension $n-3$, where $k$ is the index of $G$ in $S_1\times S_{n-1}$.
\section*{Acknowledgments}
The author would like to thank Dmitry N. Feichtner-Kozlov for this interesting problem and the support of this work, Juliane Lehmann for the helpful discussions and Giacomo d'Antonio for the hint that led to an improvement of section \ref{giacomo}.
\bibliographystyle{model1-num-names}
\begin{thebibliography}{00}
\bibitem{transfer}
G.E. Bredon, \textit{Introduction to compact transformation groups}, Pure and applied mathematics \textbf{46}, Chapter III, Academic Press, 1972.
\bibitem{emk}
E.M. Feichtner, \textit{Complexes of trees and nested set complexes}, Pacific journal of mathematics \textbf{227} (2) (2006), pp. 271-286.
\bibitem{forman}
R. Forman, \textit{Morse theory for cell complexes}, Adv. Math. \textbf{134} (1) (1998), pp. 90-145.
\bibitem{braid1}
M. Goresky, R. MacPherson, \textit{Stratified Morse theory}, Springer-Verlag, Berlin Heidelberg New York, 1988.
\bibitem{hatcher}
A. Hatcher, \textit{Algebraic Topology}, Chapter 2, Cambridge University Press, 2008.
\bibitem{clmap}
D.N. Kozlov, \textit{Closure maps on regular trisps}, Topology and its Applications \textbf{156} (15) (2009), pp. 2491-2495.
\bibitem{buch}
D.N. Kozlov, \textit{Combinatorial Algebraic Topology}, Algorithms and Computation in Mathematics \textbf{21}, Springer-Verlag Berlin Heidelberg, 2008.
\bibitem{sing1}
D.N. Kozlov, \textit{Topology of spaces of hyperbolic polynomials and combinatorics of resonances}, Israel Journal of Mathematics \textbf{132} (2002), pp. 189-206.
\bibitem{sing2}
D.N. Kozlov, \textit{Rational homology of spaces of complex monic polynomials with multiple roots}, Mathematika  \textbf{49} (2002), pp. 77-91.
\bibitem{jule}
J. Lehmann, \textit{Equivariant closure operators and trisp closure maps}, Topology and its Applications \textbf{157} (7) (2010), pp. 1195-1201.
\bibitem{braid2}
P. Orlik, H. Terao, \textit{Arrangements of Hyperplanes}, A Series of Comprehensive Studies in Mathematics \textbf{300}, Springer-Verlag, 1992.
\bibitem{white}
A. Robinson, S. Whitehouse, \textit{The tree representation of $\Sigma_{n+1}$}, Journal of Pure and Applied Algebra \textbf{111} (1996), pp. 245-253.
\bibitem{sing3}
V.A. Vassiliev, \textit{Complements of Discriminants of Smooth Maps: Topology and its Applications}, Translations of Mathematical Monographs \textbf{98}, American Mathematical Society, 1994.
\end{thebibliography}
\end{document}

>>>>

In this paper we give an answer for all Young subgroups that fix the element $1$, which are of the form $S_1\times S_{k_1}\times\dots\times S_{k_r}$, as a corollary of the more general Proposition \ref{rdthm}. An instance, which we have not yet solved completely, are Young subgroups of the form $S_k\times S_{n-k}$. However, we have computed the first homology with integer coefficients, obtaining the result $H_1(\Delta(\overline{\Pi}_n)/S_k\times S_{n-k};\Z)\cong 0$ for $n\geq 5$.
\section{Discrete Morse Theory}
The definitions of regular trisps, partial matchings, acyclic matchings and foundations of Discrete Morse Theory can be found in \cite{forman}, \cite{clmap} and \cite{buch}. The following two theorems of Discrete Morse Theory are frequently used in our proofs. We also make use of the results in \cite{clmap}.
\begin{thm}[Patchwork Theorem]
\label{patchwork}
Let $\varphi:P\longrightarrow Q$ be an order-preserving map and assume we have acyclic matchings on the subposets $\varphi^{-1}(q)$ for all $q\in Q$. Then the union of these matchings is an acyclic matching on $P$.
\end{thm}
\begin{thm}%[Main Theorem of Discrete Morse Theory for CW Complexes]
\label{morsemain}
Let $\Delta$ be a regular trisp and let $M$ be an acyclic matching on the poset ${\cal F}(\Delta)\setminus\{\hat{0}\}$. Let $c_i$ denote the number of critical $i$-dimensional simplices of $\Delta$. Then:
\begin{enumerate}
\item If the critical simplices form a subcomplex $\Delta_c$ of $\Delta$, then $\Delta$ collapses onto $\Delta_c$.
\item $\Delta$ is homotopy equivalent to a CW complex with $c_i$ cells of dimension $i$.
\end{enumerate}
\end{thm}
The proofs of Theorems \ref{patchwork} and \ref{morsemain} as well as further facts on Discrete Morse Theory can be found in \cite[Chapter 11]{buch}.
\section{The collapsible subtrisp}
\label{allgemein}
Let $n\geq 3$ be a fixed natural number throughout this section. The symmetric group $S_n$ operates on $\Delta(\overline{\Pi}_n)$ in a natural way. In this paper we consider subgroups of the Young subgroup
\[S_1\times S_{n-1}:=\{\sigma\in S_n\mid\sigma(1)=1\}\]
We now consider the regular trisp $\Delta:=\Delta(\overline{\Pi}_n)/G$, where $G\subset S_1\times S_{n-1}$ is a subgroup. Let $\Delta^*$ be that subtrisp of $\Delta$ that we obtain by removing all vertices that are represented by some partition where the block containing $1$ has exactly two elements and any other block is singleton. This condition is independent of the choice of the representative. Such a representative can be written as
\[
v_k:=\{\{1,k\},\{2\},\dots,\widehat{\{k\}},\dots,\{n\}\}
\]
with $k\in\{2,\ldots, n\}$. The element with the hat above is omitted.

@@

Let $B$ be the set of all vertices that are not contained in $R$. Now we define a closure map $\varphi:B\longrightarrow R$. Let $b\in B$ and let $\tilde b\in\overline{\Pi}_n$ be a representative of $b$. We modify the blocks in $\tilde b$ as follows. We replace the block $A$ containing $1$ by the two blocks $\{1\}$ and $A\setminus\{1\}$ and take the class of this new partition as $\varphi(b)$. By applying Corollary 2.5 in \cite{clmap} and Proposition 7 in \cite{jule} it is easy to see that this map is well-defined and a closure map\footnote{See Definition 2.1 in \cite{clmap}}. By Theorem 2.2 in \cite{clmap} we get an acyclic matching on ${\cal F}(\Delta^*)$ where a simplex $\sigma$ is critical if and only if  $V(\sigma)\subset R$.

@@

What about the remaining simplices that are represented by a chain that contains a $v_k$ with $k\in\{2,\ldots, n\}$? Let $V_G$ be the set of all these special vertices. We define a poset $P_G:=V_G\cup\{0\}$ such that $0$ is the smallest element of $P_G$ and the only element that is comparable with some other element. That means $x,y\in P_G$, $x<y$ implies $x=0$. We define an order-preserving map $\varphi_G:{\cal F}(\Delta)\longrightarrow P_G$ as follows. Let $\sigma\in{\cal F}(\Delta)$, then we map $\sigma$ to $0$ if $\sigma\in{\cal F}(\Delta^*)$. Otherwise we map $\sigma$ to the special vertex of $V_G$ that belongs to $\sigma$, which is unique. By Lemma \ref{mitte} we already have an acyclic matching on $\varphi_G^{-1}(0)$ with one critical cell which has dimension $0$. The next step would be to find acyclic matchings on $\varphi_G^{-1}(v)$ for all $v\in V_G$ and apply Theorem \ref{patchwork}.

@@

as follows. Let $\sigma\in{\cal F}(\Delta(\overline{\Pi}_{n-1})/i_k^{-1}(G))$ be a simplex, then we choose a representative $\tilde\sigma\in{\cal F}(\Delta(\overline{\Pi}_{n-1}))$ and describe what happens to the vertices of $\tilde\sigma$, which are partitions of the set $[n-1]$. Let $p=\{B_1,B_2,\dots,B_r\}$, where $B_1$ is the block that contains $1$, be a partition in the chain $\tilde\sigma$. For $p$ we define $p^*:=\{f_k(B_1)\cup\{k\},f_k(B_2),\dots,f_k(B_r)\}$, which is a partition of the set $[n]$. $\tilde\sigma$ can be written as

@@

By the discussion at the end of Section \ref{allgemein} it remains to show that there exists an acyclic matching on $\varphi_G^{-1}(v)$ for each $v\in V_G$ such that all critical simplices have dimension $n-3$. Let $v\in V_G$, then we choose a $2\leq k\leq n$ such that $v_k$ is a representative of $v$. By induction hypothesis we have an acyclic matching on $\Delta(\overline{\Pi}_{n-1})/i_k^{-1}(G)$ which has one critical simplex of dimension $0$ and critical simplices of dimension $n-4$. We apply Lemma \ref{indlemma} and get our acyclic matching on $\varphi_G^{-1}(v)$.
\end{proof}
\section{Counting spheres}
\label{giacomo}
We are now interested in the number of spheres we have. This can be computed either by looking into the proof of Proposition \ref{rdthm} or by determining the Betti Number $\beta_{n-3}(\Delta(\overline{\Pi}_n)/G)$, which is the dimension of $H_{n-3}(\Delta(\overline{\Pi}_n)/G;\C)$.

The action of $S_1\times S_{n-1}$ on $\Delta(\overline{\Pi}_n)$ induces an action on $H_{n-3}(\Delta(\overline{\Pi}_n);\C)$. By applying Corollary 3.3 in \cite{emk} and Theorem 1.5 in \cite{white} we know that $H_{n-3}(\Delta(\overline{\Pi}_n);\C)$ and
\[\C\cdot(S_1\times S_{n-1})=\bigoplus_{\sigma\in S_1\times S_{n-1}}\C\cdot\sigma\]
are isomorphic as $(S_1\times S_{n-1})$-modules. The Transfer Theorem\footnote{See \cite{transfer}} tells us
\[
H_{n-3}(\Delta(\overline{\Pi}_n)/G;\C)\cong H_{n-3}(\Delta(\overline{\Pi}_n);\C)^G\cong(\C\cdot(S_1\times S_{n-1}))^G
\]
The dimension of $(\C\cdot(S_1\times S_{n-1}))^G$ is the index of $G$ in $S_1\times S_{n-1}$. Hence we obtain the following result: $\Delta(\overline{\Pi}_n)/G$ is homotopy equivalent to a wedge of $k$ spheres of dimension $n-3$, where $k$ is the index of $G$ in $S_1\times S_{n-1}$.
\section*{Acknowledgments}
The author would like to thank Dmitry N. Feichtner-Kozlov for this interesting problem and the support of this work, Juliane Lehmann for the helpful discussions and Giacomo d'Antonio for the hint that led to an improvement of Section \ref{giacomo}.
\bibliographystyle{model1-num-names}
\begin{thebibliography}{00}
\bibitem{transfer}
G.E. Bredon, \textit{Introduction to compact transformation groups}, Pure and applied mathematics \textbf{46}, Chapter III, Academic Press, 1972.
\bibitem{emk}
E.M. Feichtner, \textit{Complexes of trees and nested set complexes}, Pacific journal of mathematics \textbf{227} (2) (2006), pp. 271-286.
\bibitem{forman}
R. Forman, \textit{Morse theory for cell complexes}, Adv. Math. \textbf{134} (1) (1998), pp. 90-145.
\bibitem{braid1}
M. Goresky, R. MacPherson, \textit{Stratified Morse theory}, Springer-Verlag, Berlin Heidelberg New York, 1988.
\bibitem{hatcher}
A. Hatcher, \textit{Algebraic Topology}, Chapter 2, Cambridge University Press, 2008.
\bibitem{clmap}
D.N. Kozlov, \textit{Closure maps on regular trisps}, Topology and its Applications \textbf{156} (15) (2009), pp. 2491-2495.
\bibitem{buch}
D.N. Kozlov, \textit{Combinatorial Algebraic Topology}, Algorithms and Computation in Mathematics \textbf{21}, Springer-Verlag Berlin Heidelberg, 2008.
\bibitem{sing1}
D.N. Kozlov, \textit{Topology of spaces of hyperbolic polynomials and combinatorics of resonances}, Israel Journal of Mathematics \textbf{132} (2002), pp. 189-206.
\bibitem{sing2}
D.N. Kozlov, \textit{Rational homology of spaces of complex monic polynomials with multiple roots}, Mathematika  \textbf{49} (2002), pp. 77-91.
\bibitem{jule}
J. Lehmann, \textit{Equivariant closure operators and trisp closure maps}, Topology and its Applications \textbf{157} (7) (2010), pp. 1195-1201.
\bibitem{braid2}
P. Orlik, H. Terao, \textit{Arrangements of Hyperplanes}, A Series of Comprehensive Studies in Mathematics \textbf{300}, Springer-Verlag, 1992.
\bibitem{white}
A. Robinson, S. Whitehouse, \textit{The tree representation of $\Sigma_{n+1}$}, Journal of Pure and Applied Algebra \textbf{111} (1996), pp. 245-253.
\bibitem{sing3}
V.A. Vassiliev, \textit{Complements of Discriminants of Smooth Maps: Topology and its Applications}, Translations of Mathematical Monographs \textbf{98}, American Mathematical Society, 1994.
\end{thebibliography}
\end{document}
